Supplementary Materials for CBEA: Competitive Balances for Taxonomic Enrichment Analysis

### 2022-01-27 cbea_revised.tex

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/data_ss_pwr_new.png}
    \caption{Statistical power (\textbf{A}) and score rankings (\textbf{B}) to assess phenotype relevance. Power (\emph{x}-axis) was estimated as the overall fraction of aerobic microbes found to be enriched in supragingival samples at $\alpha = 0.05$. 95\% confidence intervals were computed using the Agresti-Couli approach for binomial proportions. Score rankings were evaluated by comparing computed scores against true values using AUCROC (\emph{x}-axis). DeLong 95 \% confidence intervals for AUROC were computed.} 
    \label{fig:3}
\end{figure}

@@

We also evaluate phenotype relevance by assessing whether enriched sets according to ground truth are preferentially ranked higher using CBEA scores. This aspect is captured through computing AUROC by comparing computed enrichment scores and true labels. Consistent with prior approaches, adjusting for correlation did not improve performance, where the AUROC values are equivalent to using the Wilcoxon Rank sum statistic at around 0.5. Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values are lower than alternate single sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.  
Unfortunately, evaluating power and score rankings using real data does not allow for observing performance trends when effect sizes, sparsity, and inter-taxa correlation are varied. In equivalent simulation analyses, we observed the opposite effect, where

\subsection*{Differential abundance analysis}
CBEA generates sample-specific scores representing the degree of enrichment of a pre-defined set. As such, we want to assess the ability to use these scores for differential abundance analysis in combination with a standard difference of means statistical test (Welch's t-test and Wilcoxon rank sum test). We compared the performance of this approach with CBEA and two commonly used methods for differential abundance testing in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.

\subsubsection*{Simulation studies}
\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/sim_diff_ab_comb.png}
    \caption{Type I error rate \textbf{(A)} and power \textbf{(B)} for differential abundance test across different parametric simulation scenarios. For CBEA methods, differential abundance analysis was performed using a difference in means test (either Wilcoxon rank-sum test or Welch's t-test) across case/control status using single sample scores generated by CBEA (across different output types and distributional assumptions). CBEA associated methods demonstrated similar type I error to conventional differential abundance analysis methods but with more power to detect differences even at small effect sizes.} 
    \label{fig:4}
\end{figure}

Fig.~\ref{fig:4} present results for simulation studies for both type I error (panel A) and power (panel B) evaluations. All methods control for type I error well across both sparsity and correlation levels, where the estimated rate was consistently around the 0.05 pre-defined threshold. Results were similar across all evaluated methods, although in some instances, for example in medium correlation setting ($\rho = 0.2$), the unadjusted CBEA resulted in higher type I error, regardless of difference in means test and distribution of choice.

The difference between the methods is more noticeable when evaluating power. All CBEA associated variants showed much higher power even when the effect size is limited (fold change is 1.5), and there is a noticeable gap in performance between CBEA and both DESeq2 and corncob. Surprisingly, this effect is consistent across correlation levels and sparsity, even though we expectedly see performance in power drop as a function of sparsity especially in low effect size settings.

\subsubsection*{Real data evaluations}
In addition to simulation studies, we also evaluated performance of the methods on real 16S rRNA gene sequencing data set from HMP (Fig~\ref{fig:5}). For type I error evaluations, we use stool samples and randomly assign them with case/control status and calculated type I error as the proportion of genera identified as significantly different. For true positive rate evaluations, we use the gingival data set as detailed in the previous section, and calculated the true positive rate as the proportion of genera labeled as either anaerobic or aerobic that were found to be significant.

We observed both corncob and DESeq2 had significantly inflated type I error rate while all variations of CBEA were controlling for type I error at the defined $\alpha$ threshold of 0.05. This is surprising given the consistency of preserving type I error for both corncob and DESeq2 in all simulation evaluations.

In true positive experiments with data from the gingival site, estimated rates were more similar across the different methods. As expected, using the Wilcoxon rank sum test resulted in lower true positive rate compared to remaining methods, but the difference was not noticable. This is also surprising given that in simulation studies, both corncob and DESeq2 showed markedly lower power across all effect sizes.

\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/data_diff_ab.png}
    \caption{Differential abundance analysis using corncob, DESeq2 and CBEA with either Wilcoxon rank sum test or Welch's t-test. Panel \textbf{(A)} shows type I error results as the proportion of significant genera after 500 iterations where case/control status was assigned randomly to each sample. Panel \textbf{(B)} shows true positive rate results as the proportion of significant genera who are either obligate anaerobes or aerobes. Both evaluations use 16S rRNA gene sequencing data from HMP. Type I error evaluation used stool samples while the true positive rate evaluation used samples from the gingival site. Results showed that CBEA associated methods were able to keep type I error rate at approximately 0.05 while still demonstrating similar power as both corncob and DESeq2} 
    \label{fig:5}
\end{figure}

>>>>

\subsubsection*{Inference at the population level}

\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/data_diffab_fdr.eps}
    \caption{Statistical power (\textbf{A}) and score rankings (\textbf{B}) to assess phenotype relevance. Power (\emph{x}-axis) was estimated as the overall fraction of aerobic microbes found to be enriched in supragingival samples at $\alpha = 0.05$. 95\% confidence intervals were computed using the Agresti-Couli approach for binomial proportions. Score rankings were evaluated by comparing computed scores against true values using AUCROC (\emph{x}-axis). DeLong 95 \% confidence intervals for AUROC were computed.} 
    \label{fig:3}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/data_ss_pwr_new.png}
    \caption{Statistical power (\textbf{A}) and score rankings (\textbf{B}) to assess phenotype relevance. Power (\emph{x}-axis) was estimated as the overall fraction of aerobic microbes found to be enriched in supragingival samples at $\alpha = 0.05$. 95\% confidence intervals were computed using the Agresti-Couli approach for binomial proportions. Score rankings were evaluated by comparing computed scores against true values using AUCROC (\emph{x}-axis). DeLong 95 \% confidence intervals for AUROC were computed.} 
    \label{fig:4}
\end{figure}

@@

We also evaluate phenotype relevance by assessing whether enriched sets according to ground truth are preferentially ranked higher using CBEA scores. This aspect is captured through computing AUROC by comparing computed enrichment scores and true labels. Consistent with prior approaches, adjusting for correlation did not improve performance, where the AUROC values are equivalent to using the Wilcoxon Rank sum statistic at around 0.5. Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values are lower than alternate single sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.

Unfortunately, evaluating power and score rankings using real data does not allow for observing performance trends when effect sizes, sparsity, and inter-taxa correlation are varied. In equivalent simulation analyses, we observed the opposite effect, where
